\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2021-11-26 main.Rmd
a second sketch of introduction

In spatio-temporal data analysis, there's a need for a new spato-temporal data structure. The requirement for such a tool is necessary given the ubiquity of spatio-temporal data in the wild. Example of this includes house price of a city or county, climate station data from the government. The ireland wind data from

@@

Many data structures have been proposed for spatial and temporal data, but not many for spatio-temporal data. One of the reason could be the inherent different levels of information make it inefficient to store in the same table. `spacetime`[@spacetime] proposed four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT) based on underlying spatial structure `sp` [@sp] and temporal structure `xts`[@xts]. `spatstat` [@spatstat] implements a `ppp` class for point pattern data. More recent package `stars` [@stars] uses a spatiotemporal array to store the data and the array structure has its influence from `cubelyr` [@cubelyr], a dplyr data cube backend. \newline

@@

@@

**Spatio-temporal data + examples**  
Spatio-temporal data record changes of variables in spatially separated regions across time. In this article, we consider spatio-temporal vector data, which are recorded in a fixed interval. Examples of this type of data include the house price of a city or county, climate measures from weather stations in a country, and river level data from electronic gauges \textcolor{red}{more examples here}. \newline

>>>>

Many data structures have been proposed for spatial (`sf` by @sf) and temporal (`tsibble` by @tsibble) data in the R community, while less has been done for spatio-temporal data. The lack of such tools could potentially  because analysts usually treat the spatial and temporal dimension separately, without realising the need to create a new data structure for spatio-temporal data. While this approach fulfills the third tidy data principal to store each type of observational unit in its own table [@tidydata], analysts will always need to manually join the station summary of temporal results into the spatial table or combine the two tables into one for downstream analysis. This additional step doesn't add new operations but can be error prone.  \newline

@@

Currently, available spatio-temporal data structure in R includes: `spacetime`[@spacetime], which has proposed four space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI), and trajectory (STT). The data structure it uses are based on `sp` [@sp] for spatial and `xts`[@xts] for temporal, both of which has been replaced by more recent implementations. `spatstat` [@spatstat] implements a `ppp` class for point pattern data. More recent package `stars` [@stars] uses a spatio-temporal array to store the data and the array structure has its influence from `cubelyr` [@cubelyr], a dplyr data cube backend. None of these packages provide [....] and this motivates a new spatio-temporal data structure that can smoothly switch between the spatial and temporal table for manipulation of each dimension, while also take care of combining the two dimension together. The requirement for such a tool is important given the ubiquity of spatio-temporal vector data in the wild. Example of this includes [house price of a city or county], climate station data from Beureau of Meteorology (BoM), the Ireland wind data from `gstat`, air pollution data. \newline

@@

##################################################################

@@

 * In this article, we consider spatio-temporal vector data, which are recorded in a fixed interval.

### 2021-11-26 main.tex
a second sketch of introduction

In spatio-temporal data analysis, there's a need for a new
spato-temporal data structure. The requirement for such a tool is
necessary given the ubiquity of spatio-temporal data in the wild.
Example of this includes house price of a city or county, climate
station data from the government,

@@

Many data structures have been proposed for spatial and temporal data,
but not many for spatio-temporal data. One of the reason could be the
inherent different levels of information make it inefficient to store in
the same table. \texttt{spacetime}\citep{spacetime} proposed four
space-time layouts: Full grid (STF), sparse grid(STS), irregular (STI),
and trajectory (STT) based on underlying spatial structure \texttt{sp}
\citep{sp} and temporal structure \texttt{xts}\citep{xts}.
\texttt{spatstat} \citep{spatstat} implements a \texttt{ppp} class for
point pattern data. More recent package \texttt{stars} \citep{stars}
uses a spatiotemporal array to store the data and the array structure
has its influence from \texttt{cubelyr} \citep{cubelyr}, a dplyr data
cube backend. \newline

@@

\textbf{Spatio-temporal data + examples}\\
Spatio-temporal data record changes of variables in spatially separated
regions across time. In this article, we consider spatio-temporal vector
data, which are recorded in a fixed interval. Examples of this type of
data include the house price of a city or county, climate measures from
weather stations in a country, and river level data from electronic
gauges \textcolor{red}{more examples here}. \newline

>>>>

Many data structures have been proposed for spatial (\texttt{sf} by
\citet{sf}) and temporal (\texttt{tsibble} by \citet{tsibble}) data in
the R community, while less has been done for spatio-temporal data. The
lack of such tools could potentially because analysts usually treat the
spatial and temporal dimension separately, without realising the need to
create a new data structure for spatio-temporal data. While this
approach fulfills the third tidy data principal to store each type of
observational unit in its own table \citep{tidydata}, analysts will
always need to manually join the station summary of temporal results
into the spatial table or combine the two tables into one for downstream
analysis. This additional step doesn't add new operations but can be
error prone. \newline

@@

Currently, available spatio-temporal data structure in R includes:
\texttt{spacetime}\citep{spacetime}, which has proposed four space-time
layouts: Full grid (STF), sparse grid(STS), irregular (STI), and
trajectory (STT). The data structure it uses are based on \texttt{sp}
\citep{sp} for spatial and \texttt{xts}\citep{xts} for temporal, both of
which has been replaced by more recent implementations.
\texttt{spatstat} \citep{spatstat} implements a \texttt{ppp} class for
point pattern data. More recent package \texttt{stars} \citep{stars}
uses a spatio-temporal array to store the data and the array structure
has its influence from \texttt{cubelyr} \citep{cubelyr}, a dplyr data
cube backend. None of these packages provide {[}\ldots.{]} and this
motivates a new spatio-temporal data structure that can smoothly switch
between the spatial and temporal table for manipulation of each
dimension, while also take care of combining the two dimension together.
The requirement for such a tool is important given the ubiquity of
spatio-temporal vector data in the wild. Example of this includes
{[}house price of a city or county{]}, climate station data from Beureau
of Meteorology (BoM), the Ireland wind data from \texttt{gstat}, air
pollution data. \newline

@@

\begin{itemize}
\tightlist
\item
  In this article, we consider spatio-temporal vector data, which are
  recorded in a fixed interval.
\end{itemize}
